A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2017-01-06 imaging_demo/ms.tex
Add keyword 'telescopes', remove unused comments
From 01314a1 (mnras):
====================

In ms.tex,

(*) Add 'telescopes' to list of keywords
(*) Remove unused comments
Emphasize heterogeneous array imaging framework
From e50a08f (mnras):
====================

In ms.tex,

(*) Emphasize the extra message that a mathematical framework is
provided for imaging with heterogeneous arrays in abstract, intro
and conclusions
(*) Remove erroneous superscript in the estimated weights in equation
B15 in appendix B
(*) Strikethrough the sentence on "taking advantage of the
multiplication-convolution th

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage[flushleft]{threeparttable} % For table notes
\usepackage[caption=false]{subfig} % For subfloats

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes advantage of the multiplication-convolution theorem of Fourier transforms. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular, parallelizable, implemented in object-oriented Python, and publicly available. We have verified the images produced to be equivalent to those from traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). We show that it robustly estimates the input sky model for heterogeneous arrays. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying an optimized version of EPIC. The algorithm is a strong candidate for instruments targeting transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena due to the availability of high-speed calibrated time-domain images and low output bandwidth relative to visibility-based systems.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
instrumentation: interferometers -- techniques: image processing -- techniques: interferometric
\end{keywords}

@@

This paradigm shift requires a fundamentally new approach to the design of
digital correlators \citep{lon00}. Modern correlators calculate the cross-power
correlation between all antenna pairs in many narrow frequencies, forming
\emph{visibilities}, the fundamental measurement of traditional radio
interferometers. The computational requirements for a modern FX correlator scale
with the number of antenna pairs, or the square of the number of antennas $\sim
\Nant^2$ \citep{bun04}. For this reason traditional correlators have difficulty
scaling to thousands of antennas. As an example, the full HERA correlator for
352 dishes with 200 MHz of bandwidth requires 212 trillion complex multiplies
and adds per second (TMACS). Future arrays with thousands of collecting elements
will require orders of magnitude more computation, making the correlator the
dominant cost.

@@

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. EPIC is a highly parallel, object oriented Python package that primarily 
implements the MOFF imaging algorithm as well as emulating real-life telescopes and 
FX/XF correlators in software, and includes a visibility-based imaging technique 
for reference. It is intended to provide a development platform to test different 
imaging approaches, characterize scaling relations and serve as a stepping stone 
for real-life GPU/FPGA-based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. \S\ref{sec:versatility} demonstrates its ability to robustly image data from heterogeneous arrays. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator is computationally more efficient, and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Electric fields from astrophysical sources, $\mathcal{E}(\hat{\mathbf{s}}, f, t)$, in the sky coordinate system denoted by the unit vector $\hat{\mathbf{s}}$, propagate towards the observer as:
\begin{align}
  E(\mathbf{r}, f, t) &= \int \mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}\cdot\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, $\mathbf{r}$ denotes the observer's location, $f$ is the frequency of radiation, $c$ is the speed of light, $t$ denotes time, $\dif\Omega$ is the infinitesimal solid angle element on the celestial sphere whose normal is the unit vector $\hat{\mathbf{s}}$, and $E(\mathbf{r}, f, t)$ is the propagated electric field. Thus the propagated electric field is a linear superposition of the electric fields emanating from astronomical sources with appropriate complex phases. Ignoring wide-field effects, it can be simplified as a Fourier transform of the electric fields in the sky coordinates.

An antenna, located at $\mathbf{r}_a$ (indexed by $a$), measures a phased sum of these propagated electric fields over its effective collecting area with an additive receiver noise:
\begin{align}\label{eqn:measured-E-field}
  E_a(f,t) &= \int W_a(\mathbf{r}-\mathbf{r}_a,f,t)\,E(\mathbf{r},f,t)\,\dif^2\mathbf{r} + n_a(f,t) \\
           &= \int W_a(\mathbf{r}-\mathbf{r}_a,f,t) \nonumber \\
           &\quad \times \left[ \int \mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}\cdot\hat{\mathbf{s}}/c}\,\dif\Omega \right] \dif^2\mathbf{r} + n_a(f,t) \\
           &= \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega + n_a(f,t)
\end{align}
where, $W_a(\mathbf{r},f,t)$ is the aperture electric field illumination pattern of the antenna and its Fourier transform, $\mathcal{W}_a(\hat{\mathbf{s}},f,t)$, is the directional antenna voltage response at a given frequency and time.

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. The noise-like cross-terms ideally have zero mean. This expression can be further simplified using the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)$, as $\left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t = \mathcal{I}(\hat{\mathbf{s}},f,t)\,\delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{B}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to the measurement plane, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  V_{ab}(f,t) &= \int B_{ab}(\mathbf{r}_{ab}-\mathbf{r},f,t) \nonumber\\ 
              &\qquad \times \left[\int I(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}.\hat{\mathbf{s}}/c}\,\dif\Omega\right]\dif^2\mathbf{r}\,+\, n_{ab}(f,t),
\end{align}
where, $B_{ab}(\mathbf{r},f,t)$ denotes the power response of the antenna pair obtained in the measurement plane by a spatial Fourier transform of $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$. Effectively, the multiplication in image space by $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ has been replaced by a convolution with $B_{ab}(\mathbf{r},f,t)$ in the measurement plane. This is the software holographic equivalent of a traditional FX correlator output. For context, the term ``holography'' is derived from holographic measurements of the complex antenna aperture illumination pattern \citep{ben76,sco77}. In the present context, ``software holography'' refers to the application of this antenna aperture illumination pattern in software for imaging.

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility-based imaging 
techniques. Time samples of electric fields measured by the antenna and 
digitized by the A/D converter is Fourier transformed to generate electric 
field spectra. This step can be parallelized across antennas%  as shown in 
% Fig.~\ref{fig:f-engine}
. The output is then fed to either MOFF or 
visibility-based imaging pipelines.

% \begin{figure}
%   \includegraphics[width=\columnwidth]{figure4}
%   % \includegraphics[width=\columnwidth]{parallel_F_engine}
%   % \includegraphics[width=\columnwidth]{parallel_F_engine.eps}
%   \caption{Block diagram of a F-engine. The electric field data streams from
%     antennas are Fourier transformed in parallel to generate electric field
%     spectra.}
%   \label{fig:f-engine}
% \end{figure}

@@

{\bf Heterogeneity in antenna arrays occurs primarily in two scenarios. The first is when the array is intended to be identical but differences in aperture and electrical properties arise due to deviations during the manufacturing process or from random deviant behaviour such as differences in gains between dipoles in a phased array. % In practice, due to this reason almost all arrays are heterogeneous albeit by varying degrees. 
  The second scenario is when the array is designed to be heterogeneous to gain specific advantages such as sampling a wider range of spatial frequencies and decoupling the sky brightness distribution from the antenna power pattern. This has motivated a number of heterogeneous arrays such as the Combined Array for Research in Millimeter-wave Astronomy (CARMA; \citealt{woo04,wri10}), the Atacama Large Millimeter/submillimeter Array (ALMA; \citealt{igu09}), and the Very Long Baseline Array (VLBA). The following analysis applies to both of the aforementioned scenarios although for purposes of clear demonstration, we will choose an example with large differences across antennas, making it closer to the second scenario.}

@@

The sky response of the weighted visibility, referred to as {\it fringes}, is:
\begin{align}
  \mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) &= \mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\,\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}.
\end{align}
Since all quantities in the celestial plane (calligraphic fonts) are functions of position, $\hat{\mathbf{s}}$, and frequency, $f$, we drop writing this dependence explicitly hereafter.

@@

This demonstrates that images from EPIC can robustly image data from heterogeneous arrays while being mathematically optimal. Besides being sub-optimal, erroneous assumptions about the homogeneity of an array can lead to significant and systematic mis-estimates ({\bf an underestimate in the above example}) of the sky model.

% {\bf In the above example, we chose a scenario wherein the antennas in the array have large differences between them to clearly demonstrate the functionality in EPIC to image data from heterogeneous arrays. The analysis, however, is applicable to any heterogeneous array. In cases where the differences are smaller, the fidelity with which the sky model is recovered will still be degraded, although to a correspondingly smaller extent. To robustly quantify this degradation as a function of deviation from homogeneity will require image deconvolution and will be investigated in the future as it is currently beyond the scope of this paper.}

@@

Of all steps in the MOFF pipeline that are repeated for every writeout from 
the F-engine, the most expensive step even for dense HERA layouts is found to be 
the spatial two-dimensional FFT in the imaging stage relative to applying the 
sparse matrix gridding convolution, squaring or time-averaging. For instance, 
even in the conservative dense array layout scenario that makes these other 
stages even more expensive, the gridding convolution, squaring and time-averaging 
take up only $\lesssim 20$\%, $\lesssim 20$\% and $\lesssim 5$\% respectively 
of the total processing time while the spatial Fourier transform takes up 
$\gtrsim 55$ per cent of the total time. With sparser arrays the gridding process 
will be be even faster.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure11}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of
    number of complex {\bf multiplications and additions}. The solid line
    is the boundary at which the number of operations with MOFF and 
    visibility-based imaging are equal. MOFF imaging is more efficient for 
    telescopes occupying the left of this line and vice versa. CHIME, HIRAX, 
    SKA1-LC, SKA1-LCD and all the HERA layouts except HERA-19 and HERA-37 lie 
    in the parameter space favoured by MOFF imaging. The dashed line shows the 
    projected trajectory of hypothetical expanded HERA layouts. The dotted line 
    similarly shows hypothetical expanded MWA layouts with more tiles added in 
    the same core. The grey shaded area denotes the projected trajectory of the 
    LWA bounded by LWA1 (left edge), LWA-OV (right edge), current layout 
    (bottom) and a four-fold increase in the number of elements within a 50 per 
    cent increase in the core size (top). Current instruments such as MWA and 
    LOFAR fall in a region favoured by visibility-based imaging.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

@@

As radio astronomy is entering a new era, advances in instrumentation have to be accompanied by equal advances in processing techniques to manage computational resources. Many future radio telescopes such as the SKA, HERA and LWA are headed towards the large-N dense array layout model for which computational cost from traditional FX/XF correlator-based architecture and visibility-based imaging starts rising steeply. We have provided the first software demonstration of a general purpose imaging algorithm using our generic and efficient EPIC software that is designed to bring this cost down from $\mathcal{O}(\Nant^2)$ to $\mathcal{O}(\Ngrid\log_2 \Ngrid)$. Under the class of direct imaging techniques, ours is one of the most generic -- neither does it place any constraint on the array layout to be on a regular grid nor does it require the antenna array to be homogeneous. We have demonstrated its natural ability to robustly image data from heterogeneous arrays and shown that wrong assumptions about array homogeneity could result in significant and systematic mis-estimation of the sky model.

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage[flushleft]{threeparttable} % For table notes
\usepackage[caption=false]{subfig} % For subfloats
\usepackage[normalem]{ulem} % For strikethrough text

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. \sout{It takes advantage of the multiplication-convolution theorem of Fourier transforms.} Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular, parallelizable, implemented in object-oriented Python, and publicly available. We have verified the images produced to be equivalent to those from traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). {\bf We provide a detailed framework for imaging with heterogeneous arrays and show that EPIC} robustly estimates the input sky model for {\bf such} arrays. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying an optimized version of EPIC. The algorithm is a strong candidate for instruments targeting transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena due to the availability of high-speed calibrated time-domain images and low output bandwidth relative to visibility-based systems.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
instrumentation: interferometers -- techniques: image processing -- techniques: interferometric -- telescopes
\end{keywords}

@@

This paradigm shift requires a fundamentally new approach to the design of digital correlators \citep{lon00}. Modern correlators calculate the cross-power correlation between all antenna pairs in many narrow frequencies, forming \emph{visibilities}, the fundamental measurement of traditional radio interferometers. The computational requirements for a modern FX correlator scale with the number of antenna pairs, or the square of the number of antennas $\sim \Nant^2$ \citep{bun04}. For this reason traditional correlators have difficulty scaling to thousands of antennas. As an example, the full HERA correlator for 352 dishes with 200 MHz of bandwidth requires 212 trillion complex multiply-accumulates per second (TMACS). Future arrays with thousands of collecting elements will require orders of magnitude more computation, making the correlator the dominant cost.

@@

Here we present the first software implementation of the MOFF correlator, and announce the public release of the E-field Parallel Imaging Correlator (EPIC) code. EPIC is a highly parallel, object-oriented Python package that primarily implements the MOFF imaging algorithm, \sout{as well as emulating} {\bf emulates} real-life telescopes and FX/XF correlators in software, and includes a visibility-based imaging technique for reference. It is intended to provide a development platform to test different imaging approaches, characterize scaling relations and serve as a stepping stone for real-life GPU/FPGA-based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. {\bf In \S\ref{sec:versatility}, we present a mathematical framework for imaging with heterogeneous arrays and demonstrate the ability of EPIC to robustly image data from such arrays}. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator is computationally more efficient, and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Electric fields from astrophysical sources, $\mathcal{E}(\hat{\mathbf{s}}, f, t)$, in the sky coordinate system denoted by the unit vector $\hat{\mathbf{s}}$, propagate towards the observer as:
\begin{align}
  E(\mathbf{r}, f, t) &= \int \mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}\cdot\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, $i=\sqrt{-1}$, $\mathbf{r}$ denotes the observer's location, $f$ is the frequency of radiation, $c$ is the speed of light, $t$ denotes time, $\dif\Omega$ is the infinitesimal solid angle element on the celestial sphere whose normal is the unit vector $\hat{\mathbf{s}}$, and $E(\mathbf{r}, f, t)$ is the propagated electric field. Thus the propagated electric field is a linear superposition of the electric fields emanating from astronomical sources with appropriate complex phases. Ignoring wide-field effects, it can be simplified as a Fourier transform of the electric fields in the sky coordinates.

An antenna, located at $\mathbf{r}_a$ (indexed by $a$), measures a phased sum of these propagated electric fields over its effective collecting area with an additive receiver noise:
\begin{align}\label{eqn:measured-E-field}
  E_a(f,t) &= \int W_a(\mathbf{r}-\mathbf{r}_a,f,t)\,E(\mathbf{r},f,t)\,\dif^2\mathbf{r} + n_a(f,t) \\
           &= \int W_a(\mathbf{r}-\mathbf{r}_a,f,t) \nonumber \\
           &\quad \times \left[ \int \mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}\cdot\hat{\mathbf{s}}/c}\,\dif\Omega \right] \dif^2\mathbf{r} + n_a(f,t) \\
           &= \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega + n_a(f,t)
\end{align}
where, $W_a(\mathbf{r},f,t)$ is the aperture electric field illumination pattern of the antenna and its Fourier transform, $\mathcal{W}_a(\hat{\mathbf{s}},f,t)$, is the directional antenna voltage response at a given frequency and time.

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{i 2\pi f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-i 2\pi f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. The noise-like cross-terms ideally have zero mean. This expression can be further simplified using the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)$, as $\left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t = \mathcal{I}(\hat{\mathbf{s}},f,t)\,\delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{B}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to the measurement plane, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  V_{ab}(f,t) &= \int B_{ab}(\mathbf{r}_{ab}-\mathbf{r},f,t) \nonumber\\ 
              &\qquad \times \left[\int I(\hat{\mathbf{s}},f,t)\,e^{-i 2\pi f\mathbf{r}.\hat{\mathbf{s}}/c}\,\dif\Omega\right]\dif^2\mathbf{r}\,+\, n_{ab}(f,t),
\end{align}
where, $B_{ab}(\mathbf{r},f,t)$ denotes the power response of the antenna pair obtained in the measurement plane by a spatial Fourier transform of $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$. Effectively, the multiplication in image space by $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ has been replaced by a convolution with $B_{ab}(\mathbf{r},f,t)$ in the measurement plane. This is the software holographic equivalent of a traditional FX correlator output. For context, the term ``holography'' is derived from holographic measurements of the complex antenna aperture illumination pattern \citep{ben76,sco77}. In the present context, ``software holography'' refers to the application of this antenna aperture illumination pattern in software for imaging.

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility-based imaging techniques. Time samples of electric fields measured by the antenna and digitized by the A/D converter is Fourier transformed to generate electric field spectra. This step can be parallelized across antennas. The output is then fed to either MOFF or visibility-based imaging pipelines.

@@

Heterogeneity in antenna arrays occurs primarily in two scenarios. The first is when the array is intended to be identical but differences in aperture and electrical properties arise due to deviations during the manufacturing process or from random deviant behaviour such as differences in gains between dipoles in a phased array. The second scenario is when the array is designed to be heterogeneous to gain specific advantages such as sampling a wider range of spatial frequencies and decoupling the sky brightness distribution from the antenna power pattern. This has motivated a number of heterogeneous arrays such as the Combined Array for Research in Millimeter-wave Astronomy (CARMA; \citealt{woo04,wri10}), the Atacama Large Millimeter/submillimeter Array (ALMA; \citealt{igu09}), and the Very Long Baseline Array (VLBA). The following analysis applies to both of the aforementioned scenarios although for purposes of clear demonstration, we will choose an example with large differences across antennas, making it closer to the second scenario.

@@

The sky response of the weighted visibility, referred to as {\it fringes}, is:
\begin{align}
  \mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) &= \mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\,\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\,e^{i 2\pi f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}.
\end{align}
Since all quantities in the celestial plane (calligraphic fonts) are functions of position, $\hat{\mathbf{s}}$, and frequency, $f$, we drop writing this dependence explicitly hereafter.

@@

This demonstrates that images from EPIC can robustly image data from heterogeneous arrays while being mathematically optimal. Besides being sub-optimal, erroneous assumptions about the homogeneity of an array can lead to significant and systematic mis-estimates (an underestimate in the above example) of the sky model.

@@

Of all the steps in the MOFF pipeline that are repeated for every writeout from the F-engine, the most expensive step even for dense HERA layouts is found to be the spatial two-dimensional FFT in the imaging stage relative to applying the sparse matrix gridding convolution, squaring or time-averaging. For instance, even in the conservative dense array layout scenario that makes these other stages even more expensive, the gridding convolution, squaring and time-averaging take up only $\lesssim 20$\%, $\lesssim 20$\% and $\lesssim 5$\% respectively of the total processing time while the spatial Fourier transform takes up $\gtrsim 55$ per cent of the total time. With sparser arrays the gridding process will be be even faster.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure11}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated}
  % \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of number of complex multiplications and additions. The solid line is the boundary at which the number of operations with MOFF and visibility-based imaging are equal. MOFF imaging is more efficient for telescopes occupying the left of this line and vice versa. CHIME, HIRAX, SKA1-LC, SKA1-LCD and all the HERA layouts except HERA-19 and HERA-37 lie in the parameter space favoured by MOFF imaging. The dashed line shows the projected trajectory of hypothetical expanded HERA layouts. The dotted line similarly shows hypothetical expanded MWA layouts with more tiles added in the same core. The grey shaded area denotes the projected trajectory of the LWA bounded by LWA1 (left edge), LWA-OV (right edge), current layout (bottom) and a four-fold increase in the number of elements within a 50 per cent increase in the core size (top). Current instruments such as MWA and LOFAR fall in a region favoured by visibility-based imaging.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

@@

As radio astronomy is entering a new era, advances in instrumentation have to be accompanied by equal advances in processing techniques to manage computational resources. Many future radio telescopes such as the SKA, HERA and LWA are headed towards the large-N dense array layout model for which computational cost from traditional FX/XF correlator-based architecture and visibility-based imaging starts rising steeply. We have provided the first software demonstration of a general purpose imaging algorithm using our generic and efficient EPIC software that is designed to bring this cost down from $\mathcal{O}(\Nant^2)$ to $\mathcal{O}(\Ngrid\log_2 \Ngrid)$. Under the class of direct imaging techniques, ours is one of the most generic -- neither does it place any constraint on the array layout to be on a regular grid nor does it require the antenna array to be homogeneous. {\bf We have provided a detailed mathematical framework for imaging with heterogeneous arrays, demonstrated the natural ability of EPIC to robustly image data from such arrays, and shown that wrong assumptions about array homogeneity could result in significant and systematic mis-estimation of the sky model}.

Our EPIC package, now publicly available, written in object-oriented Python, is highly modularized and parallelizable. It includes an implementation of the MOFF algorithm in addition to visibility-based software holography imaging and a data simulator for sky models. It is designed to provide a development platform to compare different imaging approaches and serve as a stepping stone for real-life GPU/FPGA-based implementation on telescopes. It has been successfully tested on simulated MWA observations as well as real LWA observations from both imaging and calibration viewpoints.
